The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-06 rational.tex
related work

\paragraph*{Declarative Continuations.} 
In his Masters thesis~\cite{Filinski:1989:DCI:648332.755574}, Filinski
proposes that continuations are a \emph{declarative} concept. He,
furthermore, introduces a symmetric extension of the $\lambda$-calculus in
which values and continuations are treated as opposites. In more detail, the
symmetric calculus contains a ``value'' fragment and a ``continuation''
fragment which are mirror images. Pairs and sums are treated as duals in the
sense that the ``value'' fragment includes pairs whose mirror image in the
``continuation'' fragment are sums. In contrast, our language includes pairs
and sums in the value fragment and two symmetries: one that maps the pairs to
fractions and another that maps the sums to subtractions.

\paragraph*{The Duality of Computation.}
The precise duality between call-by-name and call-by-value was further
investigated by Selinger~\cite{Selinger:2001:CCD:966910.966911}. Curien and
Herbelin~\cite{Curien:2000} introduce a calculus that exhibits symmetries
between values and continuations and between call-by-value and
call-by-name. The calculus includes the type $A-B$ which is the dual of
implication, i.e., a value of type $A-B$ is a context expecting a function of
type $A \rightarrow B$. Alternatively a value of type $A-B$ is also explained
as a \emph{pair} consisting of a value of type $A$ and a continuation of type
$B$. This is to be contrasted with our interpretation of a value of that type
as \emph{either} a value of $A$ or a demand for a value of type $B$. This
calculus was further analyzed and extended by
Wadler~\cite{Wadler:2003,DBLP:conf/rta/Wadler05}. The extension gives no
interpretation to the subtraction connective and like the original symmetric
calculus of Filinksi, introduces a duality that relates sums to products and
vice-versa.

\paragraph*{Subtractive Logic.} 
Rauszer~\cite{springerlink:10.1007/BF02120864,rauszer,rauszer2} introduced a
logic which contains a dual to implication. Her work has been distilled in
the form of \emph{subtractive logic}~\cite{Crolard01} which has recently been
related to coroutines~\cite{Crolard01082004} and delimited
continuations~\cite{Ariola:2009:TFD:1743339.1743381}.  In more detail,
Crolard explains the type $A-B$ as the of \emph{coroutines} with a local
environment of type $A$ and a continuation of type $B$. The description is
complicated by what is essentially the desire to enforce linearity
constraints so that coroutines cannot access the local environment of other
coroutines.

\paragraph*{Negation in Classical Linear Logic} 
Filinski~\cite{Filinski92} uses the negative types of linear logic to model
continuations. Reddy~\cite{Reddy91} generalizes this idea by interpreting the
negative types of linear logic as \emph{acceptors}, which are like
continuations in the sense that they take an input and return no
output. Acceptors however are also similar in flavor to logical variables:
they can created and instantiated later once their context of use is
determined. It is clear that, at an intuitive level, acceptors are entities
that combine elements of our negative and fractional types.

\paragraph*{The Lambek-Grishin Calculus.} The ``parsing-as-deduction'' style
of linguistic analysis uses the Lambek-Grishin calculus with the following
types: product, left division, right division, coproduct, right difference,
and left difference~\cite{Bernardi:2010:CSL:1749618.1749689}. The division
and difference types are similar to our types but because the calculus lacks
commutativity and associativity and only has limited notions of
distributivity, each connective needs a left and right version. The
Lambek-Grishin exhibits two notions of symmetry but they are unrelated to our
notions. In particular, the first notion of symmetry expresses commutativity
and the second relates products to sums and divisions to subtractions. In
contract, our two symmetries relate sums to subtractions and products to
divisions.

>>>>

The idea of ``negative types'' has appeared many times in the literature and
has often been related to some form of continuations. Fractional types are
less common but have also appeared in relation to parsing natural
languages. Although each of these previous occurrences of negative and
fractional types is somewhat related to our work, our results are
subtantially different. To clarify this point, we start by reviewing the
salient point of the major pieces of related work and conclude this section
with a summary contrasting our approach to previous work.

\paragraph*{Declarative Continuations.} 
In his Masters thesis~\cite{Filinski:1989:DCI:648332.755574}, Filinski
proposes that continuations are a \emph{declarative} concept. He,
furthermore, introduces a symmetric extension of the $\lambda$-calculus in
which call-by-value is dual to call-by-name and values are dual to
continuations. In more detail, the symmetric calculus contains a ``value''
fragment and a ``continuation'' fragment which are mirror images. Pairs and
sums are treated as duals in the sense that the ``value'' fragment includes
pairs whose mirror image in the ``continuation'' fragment are sums. In
contrast, our language includes pairs and sums in the value fragment and two
symmetries: one that maps the pairs to fractions and another that maps the
sums to subtractions.

\paragraph*{The Duality of Computation.}
The duality between call-by-name and call-by-value was further investigated
by Selinger using control
categories~\cite{Selinger:2001:CCD:966910.966911}. Curien and
Herbelin~\cite{Curien:2000} also introduce a calculus that exhibits
symmetries between values and continuations and between call-by-value and
call-by-name. The calculus includes the type $A-B$ which is the dual of
implication, i.e., a value of type $A-B$ is a context expecting a function of
type $A \rightarrow B$. Alternatively a value of type $A-B$ is also explained
as a \emph{pair} consisting of a value of type $A$ and a continuation of type
$B$. This is to be contrasted with our interpretation of a value of that type
as \emph{either} a value of $A$ or a demand for a value of type $B$. This
calculus was further analyzed and extended by
Wadler~\cite{Wadler:2003,DBLP:conf/rta/Wadler05}. The extension gives no
interpretation to the subtraction connective and like the original symmetric
calculus of Filinksi, introduces a duality that relates sums to products and
vice-versa.

\paragraph*{Subtractive Logic.} 
Rauszer~\cite{springerlink:10.1007/BF02120864,rauszer,rauszer2} introduced a
logic which contains a dual to implication. Her work has been distilled in
the form of \emph{subtractive logic}~\cite{Crolard01} which has recently been
related to coroutines~\cite{Crolard01082004} and delimited
continuations~\cite{Ariola:2009:TFD:1743339.1743381}.  In more detail,
Crolard explains the type $A-B$ as the type of \emph{coroutines} with a local
environment of type $A$ and a continuation of type $B$. The description is
complicated by what is essentially the desire to enforce linearity
constraints so that coroutines cannot access the local environment of other
coroutines.

\paragraph*{Negation in Classical Linear Logic} 
Filinski~\cite{Filinski92} uses the negative types of linear logic to model
continuations. Reddy~\cite{Reddy91} generalizes this idea by interpreting the
negative types of linear logic as \emph{acceptors}, which are like
continuations in the sense that they take an input and return no
output. Acceptors however are also similar in flavor to logical variables:
they can created and instantiated later once their context of use is
determined. Although a formal connection is lacking, it is clear that, at an
intuitive level, acceptors are entities that combine elements of our negative
and fractional types.

\paragraph*{The Lambek-Grishin Calculus.} The ``parsing-as-deduction'' style
of linguistic analysis uses the Lambek-Grishin calculus with the following
types: product, left division, right division, sum, right difference, and
left difference~\cite{Bernardi:2010:CSL:1749618.1749689}. The division and
difference types are similar to our types but because the calculus lacks
commutativity and associativity and only has limited notions of
distributivity, each connective needs a left and right version. The
Lambek-Grishin exhibits two notions of symmetry but they are unrelated to our
notions. In particular, the first notion of symmetry expresses commutativity
and the second relates products to sums and divisions to subtractions. In
contrast, our two symmetries relate sums to subtractions and products to
divisions.

\paragraph*{Our Approach.} The salient aspects of our approach are the following:
\begin{itemize}
\item Negative and fractional types have an elementary and familiar
  interpretation borrowed from the algebra of rational numbers. One can write
  any algebraic identity that is valid for the rational numbers and interpret
  it as an isomorphism with a clear computational interpretation: negative
  values flow backwards and fractional values represent fragments of
  aggregate data structures. None of the systems above has such a natural
  interpretation of negative and fractional types. 
\item Because we are \emph{not} in the context of the full
  $\lambda$-calculus, which allows arbitrary duplication and erasure of
  information, values of negative and fractional types are first-class values
  that can flow anywhere. The information-preserving computational
  infrastructure guarantees that, in a complete program, every negative
  demand will be satisfied exactly once, and every constraint imposed by a
  fractional value will also be satisfied exactly once. This property is
  shared with systems that are based on linear logic; other systems must
  impose ad hoc constraints to ensure negative values are used exactly
  once. 
\item In contrast to all the work that takes continuations as primitive
  entities of negative types, we view continuations as a derived notion that
  combines a demand for a value with constraints on how this value will be
  used to proceed with the evaluation (to the closest delimiter or to the end
  of the program). In other words, we view a continuation as a non-elementary
  notion that combines the negative types to demand a value and the
  fractional types to explain how this value will be used to continue the
  evaluation. As a consequence, the previously observed duality between
  values and continuations can be teased into two dualities: a duality
  between values flowing in one direction or the other and a duality between
  aggregate values composing and decomposing into smaller values. Arguably
  each of the dualities is more natural than a duality that maps regular
  values to a conflated notion of negative and fractional types, and hence
  requires notions like ``additive pairs'' and ``multiplicative sums.''
\end{itemize}
